Denotational Semantics of General Payment Primitives, and Its Payment System

### 2022-10-03 packages/spec-haskell/yellowpaper/Paper1.tex
chapter done! Superfluid Protocol - A Reference Implementation

This paper covers the implementation of (a).

@@

For the purpose of compositional financial contracts, it is useful to separate monetary values that
bearers can use immediately from ones that are set aside for other financial purpose. The concept
of \textit{real time balance} is thus created. \textit{Real time balance} is a
functorful\footnote{Bartosz Milewski has an excellent series on these
topics: \url{https://bartoszmilewski.com/2015/01/20/functors/}, and he uses the term ``functorful''
to convey the idea of generalized container.} of monetary values. It can be converted to a single
monetary value.

@@

\paragraph{Monetary Unit Data}

@@

    -- | ω function - apply agreement operation ~ao~ (hear: ω) to the agreement
    --                operation data ~ac~ to get a tuple of:
    --
    --   1. An updated ~ac'~.
    --
    --   2. A functorful delta of agreement monetary unit data ~mudsΔ~, which
    --      then can be monoid-appended to existing
    --      ~mudΔ~.  This is what can make an agreement scalable.
    applyAgreementOperation
        :: forall t ao aoo.
           -- indexed type aliases
           ( t ~ SFT_TS sft
           , ao ~ AgreementOperation ac
           , aoo ~ AgreementOperationOutput ac
           )
        => ac -> ao -> t -> (ac, aoo)
    data family AgreementOperation ac :: Type
    data family AgreementOperationOutputF ac :: Type -> Type
    type family AgreementOperationOutput ac = (mmuds :: Type) | mmuds -> ac
\end{code}

The $\omega$ function ($applyAgreementOperation$) is the $genPrim$ in the agreement framework. It
takes in a agreement contract, an operation onto it and the current time; then it spits out an
update of the agreement contract and a functorful of new monoidal monetary unit data.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.9\textwidth]{../assets/agreement-contract-omega-function.png}
    \caption{Agreement Contract $\omega$ Function}
    \label{fig:ac-omega}
\end{figure}

@@

It is also important to note that $AgreementContract$ itself is also a $MonetaryUnitDataClass$, but
it is not a monoid, hence it is used to replace the previous agreement contract data. This has
optimization implications, as in that if agreement contracts do not produce zero balance then they
must be included in the $\nu$ function, that can make the $\nu$ function $O(N)$ to the number of
agreement contracts a monetary unit is associated with.

It may also have been self evident that agreement contract is a analogy to the fact that it encodes
``ongoing relationships'' like legal contracts do between bearers.

@@

\subsection{Instant Transferable Agreement}

\subsection{Constant Flow Agreement}

\subsection{Distribution Agreements}

\subsection{Decaying Flow Agreement}

@@

>>>>

This paper covers the overview of the implementation of (a).

@@

For the purpose of compositional financial contracts, it is useful to separate monetary values that
bearers can use immediately (called \textit{untappedValue} in code) from ones that are set aside for
other financial purpose. The concept of \textit{real time balance} is thus created. \textit{Real
time balance} is a functorful\footnote{Bartosz Milewski has an excellent series on these
topics: \url{https://bartoszmilewski.com/2015/01/20/functors/}, and he uses the term ``functorful''
to convey the idea of generalized container.} of monetary values. It can be converted to a single
monetary value.

@@

\paragraph{Monetary Unit Data (MUD)}

@@

    -- | ω function - apply agreement operation ~ao~ (hear: ω) to the agreement
    --                operation data ~ac~ to get a tuple of:
    --
    --   1. An updated ~ac'~.
    --
    --   2. A functorful delta of agreement monetary unit data ~mudsΔ~, which
    --      then can be appended to existing ~mudΔ~.  This is what can make an
    --      agreement scalable.
    applyAgreementOperation
        :: forall t ao aoo.
           -- indexed type aliases
           ( t ~ SFT_TS sft
           , ao ~ AgreementOperation ac
           , aoo ~ AgreementOperationOutput ac
           )
        => ac -> ao -> t -> (ac, aoo)
    data family AgreementOperation ac :: Type
    data family AgreementOperationOutputF ac :: Type -> Type
    type family AgreementOperationOutput ac = (smuds :: Type) | smuds -> ac
\end{code}

The $\omega$ function ($applyAgreementOperation$) is the $genPrim$ in the agreement framework. It
takes in a agreement contract, an operation onto it and the current time; then it spits out an
update of the agreement contract and a functorful of new delta of monetary unit data.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=1\textwidth]{../assets/agreement-contract-omega-function.png}
    \caption{Agreement Contract $\omega$ Function}
    \label{fig:ac-omega}
\end{figure}

@@

It is also important to note that $AgreementContract$ itself is also a $MonetaryUnitDataClass$, but
it is not a semigroup, hence it is used to replace the previous agreement contract data. This has
optimization implications, as in that if agreement contracts do not produce zero balance then they
must be included in the $\nu$ function, that can make the $\nu$ function $O(N)$ to the number of
agreement contracts a monetary unit is associated with.

It may have been self evident that agreement contract is a analogy to the fact that it encodes
``ongoing relationships'' like legal contracts do between bearers.

\subsection{Lens Data Accessors}

In order to support \textit{index abstraction} in the denotational payment primitives, Monetary unit
data are created with \textit{lens data accessors}\footfullcite{clarke2020profunctor}:

\begin{code}
-- representation of lenses
data Lens a b s t = Lens { view :: s -> a, update :: (b, s) -> t }
-- the pro-functor version of it
type Lens s t a b = forall f. Functor f => (a -> f b) -> s -> f t
\end{code}

The pro-functor version of might seem very obscure at first, but its data representation is rather
self-explanatory: a lens is simply a pair of getter ($view$) and setter ($update$) encoded in the
data.

With the help of Lens, then we can create the payment primitives independent of the choices between
1-to-1, 1-to-N, etc. Here are how they are defined for each class of payment primitives:

\paragraph{Instant Value MUD}

This is for agreements where value is instantly transferred.

\begin{code}
class ( Default amuLs
      , SuperfluidSystemTypes sft
      ) => MonetaryUnitLenses amuLs sft | amuLs -> sft where
    untappedValue :: Lens' amuLs (UntappedValue (SFT_MVAL sft))

type MonetaryUnitData :: Type -> Type -> Type
newtype MonetaryUnitData amuLs sft =
    MkMonetaryUnitData { getMonetaryUnitLenses :: amuLs }
    deriving (Default)

instance MonetaryUnitLenses amuLs sft
    => Semigroup (MonetaryUnitData amuLs sft) where
    MkMonetaryUnitData a <> MkMonetaryUnitData b =
        let c = a & over untappedValue (+ b^.untappedValue)
        in MkMonetaryUnitData c

instance MonetaryUnitLenses amuLs sft
    => MonetaryUnitDataClass (MonetaryUnitData amuLs sft) sft where
    balanceProvided (MkMonetaryUnitData a) _ =
        typedValuesToRTB [mkAnyTypedValue $ a^.untappedValue]
\end{code}

Note that (a) $^.$ is the view function (getter) for Lens types. (b) semigroup operator $(<>)$ defines
how two monetary unit data can be combined into one.

\paragraph{Constant Flow Agreement (CFA) MUD}

A more interesting case is where monetary value is changing continuously over time at a constant
rate:

\begin{code}
class ( Default amuLs
      , SuperfluidSystemTypes sft
      ) => MonetaryUnitLenses amuLs sft | amuLs -> sft where
    settledAt    :: Lens' amuLs (SFT_TS sft)
    settledValue :: Lens' amuLs (UntappedValue (SFT_MVAL sft))
    netFlowRate  :: Lens' amuLs (SFT_MVAL sft)

type MonetaryUnitData :: Type -> Type -> Type
newtype MonetaryUnitData amuLs sft =
    MkMonetaryUnitData { getMonetaryUnitLenses :: amuLs }
    deriving (Default)

instance MonetaryUnitLenses amuLs sft
    => Semigroup (MonetaryUnitData amuLs sft) where
    MkMonetaryUnitData a <> MkMonetaryUnitData b =
        let t  = a^.settledAt
            t' = b^.settledAt
            settledΔ = MkUntappedValue $ a^.netFlowRate * fromIntegral (t' - t)
            c = a & set  settledAt    t'
                  & over netFlowRate  (+ b^.netFlowRate)
                  & over settledValue (+ (b^.settledValue + settledΔ))
        in MkMonetaryUnitData c

instance MonetaryUnitLenses amuLs sft
    => MonetaryUnitDataClass (MonetaryUnitData amuLs sft) sft where
    balanceProvided (MkMonetaryUnitData a) t =
        let b = uval_s + coerce (fr * fromIntegral (t - t_s))
        in  typedValuesToRTB [ mkAnyTypedValue b ]
        where t_s    = a^.settledAt
              uval_s = a^.settledValue
              fr     = a^.netFlowRate
\end{code}

Note the implementation of semigroup binary operator, this is where the optimization occurs: to use
$netFlowRate$ to fold monetary unit data into a single value.

\paragraph{Decaying Flow Agreement (CFA) MUD}

\begin{code}
class ( Default amuLs
      , SuperfluidSystemTypes sft
      ) => MonetaryUnitLenses amuLs sft | amuLs -> sft where
    decayingFactor  :: Lens' amuLs (SFT_FLOAT sft)
    settledAt       :: Lens' amuLs (SFT_TS sft)
    αVal            :: Lens' amuLs (SFT_FLOAT sft)
    εVal            :: Lens' amuLs (SFT_FLOAT sft)

type MonetaryUnitData :: Type -> Type -> Type
newtype MonetaryUnitData amuLs sft =
    MkMonetaryUnitData { getMonetaryUnitLenses :: amuLs }
    deriving (Default)

instance MonetaryUnitLenses amuLs sft
    => Semigroup (MonetaryUnitData amuLs sft) where
    MkMonetaryUnitData a <> MkMonetaryUnitData b =
        let c = a & set  settledAt     (b^.settledAt)
                  & over αVal          (\α -> α * exp (-λ * t_Δ) - ε')
                  & over εVal          (+ ε')
        in MkMonetaryUnitData c
        where ε'  = b^.εVal
              λ   = b^.decayingFactor
              t_Δ = fromIntegral (b^.settledAt - a^.settledAt)

instance MonetaryUnitLenses amuLs sft
    => MonetaryUnitDataClass (MonetaryUnitData amuLs sft) sft where
    balanceProvided (MkMonetaryUnitData a) t =
        let b = ceiling $ α * exp (-λ * t_Δ) + ε
        in  typedValuesToRTB [ (mkAnyTypedValue . MkUntappedValue) b ]
        where t_s   = a^.settledAt
              α     = a^.αVal
              ε     = a^.εVal
              λ     = a^.decayingFactor
              t_Δ   = fromIntegral (t - t_s)
\end{code}

\paragraph{Universal Index (UIDX)}

An universal index then can be implemented trivially by representing lenses using the record syntax
directly. For example, for constant flow monetary unit data:

\begin{code}
data MonetaryUnitLenses sft = MonetaryUnitLenses
    { settled_at    :: SFT_TS sft
    , settled_value :: UntappedValue (SFT_MVAL sft)
    , net_flow_rate :: SFT_MVAL sft
    } deriving (Generic)

deriving instance SuperfluidSystemTypes sft
    => Default (MonetaryUnitLenses sft)

-- | Monetary unit lenses for the universal index.
instance SuperfluidSystemTypes sft
    => CFMUD.MonetaryUnitLenses (MonetaryUnitLenses sft) sft where
    settledAt          = $(field 'settled_at)
    settledValue       = $(field 'settled_value)
    netFlowRate        = $(field 'net_flow_rate)
\end{code}

The $field$ is a template Haskell function to generate lens from record fields.

\paragraph{Proportional Distribution Index (PDIDX)}

Proportional distribution index describes on-going \textit{subscription} agreement contracts between
one \textit{publisher} and many \textit{subscribers} to the publisher's \textit{distribution}.

Its instant value variance is called \textit{``Instant Distribution Agreement (IDA)''}.

Its constant flow variance is called \textit{``Constant Flow Distribution Agreement (CFDA)''}.

@@

To finish, here is the illustration of data structures for the denotational payment primitives
implemented with agreement framework:

\begin{figure}[h]
    \centering
    \includegraphics[width=1\textwidth]{../assets/agreement-data-structures.png}
    \caption{Agreement Data Structures}
    \label{fig:ac-omega}
\end{figure}

Note that with \textit{Root MUD}, any monetary unit can traverse all its MUDs; and with \textit{Root
Contract} relations between monetary units can also be searched by a simple algorithm.

This concludes the overview of the Superfluid protocol implementation. For more details, one should
refer to the code repository mentioned before.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Acknowledgments}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
